---
title: "Analyzing Wisconsin District Elections"
---

```{r}
#| message: false
#| warning: false
#| echo: false
#| output: false

library(fec16)
library(maps)
library(tidyverse)
library(mdsr) 
library(viridis)
library(sf) 
library(ggspatial)
library(prettymapr)
library(ggplot2)
library(ggmap)
library(leaflet)
library(tmap)

print(results_house, width = Inf)
results_house |>
  group_by(state, district_id) |>
  summarize(N = n())
```

```{r}
#| message: false
#| warning: false
#| echo: false
#| output: false

district_elections <- results_house |>
  mutate(district = parse_number(district_id)) |>
  group_by(state, district) |>
  summarize(
    N = n(), 
    total_votes = sum(general_votes, na.rm = TRUE),
    d_votes = sum(ifelse(party == "DEM", general_votes, 0), na.rm = TRUE),
    r_votes = sum(ifelse(party == "REP", general_votes, 0), na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    other_votes = total_votes - d_votes - r_votes,
    r_prop = r_votes / total_votes,  
    winner = ifelse(r_votes > d_votes, "Republican", "Democrat")
  )

wi_results <- district_elections |>
  filter(state == "WI")
wi_results |>                  
  select(-state)
```


```{r}
#| message: false
#| warning: false
#| echo: false
#| output: false

wi_results |>
  skim(total_votes) |>
  select(-na)

wi_results |>
  summarize(
    N = n(), 
    state_votes = sum(total_votes), 
    state_d = sum(d_votes), 
    state_r = sum(r_votes)
  ) |>
  mutate(
    d_prop = state_d / state_votes, 
    r_prop = state_r / state_votes
  )

wi_results |>
  select(district, r_prop, winner, state) |>
  arrange(desc(r_prop))
```


```{r}
#| message: false
#| warning: false
#| echo: false
#| output: false

src <- "http://cdmaps.polisci.ucla.edu/shp/districts114.zip"
lcl_zip <- fs::path(tempdir(), "districts114.zip")
download.file(src, destfile = lcl_zip)
lcl_districts <- fs::path(tempdir(), "districts114")
unzip(lcl_zip, exdir = lcl_districts)
dsn_districts <- fs::path(lcl_districts, "districtShapes")

st_layers(dsn_districts)

districts <- st_read(dsn_districts, layer = "districts114") |>
  mutate(DISTRICT = parse_number(as.character(DISTRICT)))

wi_shp <- districts |>
  filter(STATENAME == "Wisconsin")
```

### Understanding the Data

This analysis delves into the data from the 2016 elections across Wisconsin's congressional districts. Using sophisticated geospatial techniques, we created a digital map that served as the canvas for our exploration. 

```{r}
#| message: false
#| warning: false
#| echo: false

wi_shp |>
  st_geometry() |>
  plot(col = gray.colors(nrow(wi_shp)))
```

### Insights 

The resulting map vividly displays the distribution of Republican and Democratic support, revealing the intricate electoral dynamics at play. By examining neighboring districts, we uncovered the subtle influences and relationships that shape the outcome of elections. This insight allows us to grasp the interconnectedness of communities and understand how it impacts the final electoral results.

```{r}
#| message: false
#| warning: false
#| echo: false

wi_merged <- wi_shp |>
  st_transform("+proj=longlat +datum=WGS84") |>
  inner_join(wi_results, by = c("DISTRICT" = "district"))

wi <- ggplot(data = wi_merged, aes(fill = winner)) +
  annotation_map_tile(zoom = 6, type = "osm", progress = "none") + 
  geom_sf(alpha = 0.5) +
  scale_fill_manual("Winner", values = c("blue", "red")) + 
  geom_sf_label(aes(label = DISTRICT), fill = "white") + 
  theme_void()

wi +
  aes(fill = r_prop) + 
  scale_fill_distiller(
    "Proportion\nRepublican", 
    palette = "RdBu", 
    limits = c(0.2, 0.8)
  )
```

### Interactive Exploration

To enhance our understanding, we utilized interactive mapping technology, enabling users to zoom in and explore individual cities within each district. This interactive experience offers a deeper insight into the unique characteristics and political landscapes of each area, enriching our comprehension of Wisconsin's diverse electorate.

```{r}
#| message: false
#| warning: false
#| echo: false

library(leaflet)
pal <- colorNumeric(palette = "RdBu", domain = c(0, 1))

leaflet_wi <- leaflet(wi_merged) |>
  addTiles() |>
  addPolygons(
    weight = 1, fillOpacity = 0.7, 
    color = ~pal(1 - r_prop), 
    popup = ~paste("District", DISTRICT, "</br>", round(r_prop, 4))
  ) |>                        
  setView(lng = -89.5, lat = 44.5, zoom = 6.4)
leaflet_wi
```

### Conclusion: A Glimpse into Democracy

In conclusion, our analysis provides more than just a glimpse into Wisconsin's electoral landscape; it offers a profound insight into the essence of democracy itself. From the mosaic of opinions to the spirit of civic engagement, every aspect of our analysis underscores the fundamental principles of democracy. Moreover, by shedding light on the intricacies of districting and its influence on electoral outcomes, we gain a deeper appreciation for the complexities inherent in the democratic process.
